---
title: "Getting Started with ratiod"
author: "Gilles Colling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with ratiod}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = FALSE
)
```

## Installation

ratiod requires CmdStan for Bayesian inference. Install in order:

```{r install, eval = FALSE}
# 1. Install cmdstanr (Stan interface)
install.packages("cmdstanr", repos = c("https://stan-dev.r-universe.dev", getOption("repos")))

# 2. Install CmdStan (one-time setup)
cmdstanr::install_cmdstan()

# 3. Install ratiod
# install.packages("pak")
pak::pak("gcol33/ratiod")
```

Verify installation:

```{r verify, eval = FALSE}
library(ratiod)
cmdstanr::cmdstan_version()
```

## Basic Usage

### Example: CPUE Analysis

Catch per unit effort with site-level variation:

```{r cpue-example}
library(ratiod)

# Simulated trawl data
set.seed(123)
n_sites <- 20
n_obs <- 200

trawl_data <- data.frame(
  site = factor(rep(1:n_sites, each = n_obs/n_sites)),
  depth = rnorm(n_obs, 50, 15),
  season = factor(sample(c("spring", "summer", "fall"), n_obs, replace = TRUE)),
  catch = rpois(n_obs, lambda = 15),
  effort = rgamma(n_obs, shape = 5, rate = 1)
)

# Fit joint model using combined formula syntax
fit <- ratiod(
  catch | effort ~ depth + season + (1 | site),
  data = trawl_data,
  family = ratiod_poisson_gamma()
)

# View results
print(fit)
summary(fit)
```

### Extract CPUE Posteriors

```{r cpue-extract}
# Observation-level CPUE
cpue <- ratio(fit)
summary(cpue)

# Site-level CPUE (averaged within site)
cpue_site <- ratio(fit, by = "site")
summary(cpue_site)

# Compare CPUE between seasons
ratio_contrast(fit, ~ season)
```

## The Three Families

### Count/Count Ratios: `ratiod_negbin_negbin()`

When both numerator and denominator are counts:

```{r negbin-example}
# Relative abundance: species count / total count
fit_abundance <- ratiod(
  species_count | total_count ~ habitat + (1 | plot),
  data = survey_data,
  family = ratiod_negbin_negbin()
)
```

### Trial-Based Proportions: `ratiod_binomial()`

When numerator is successes out of denominator trials:

```{r binomial-example}
# Detection probability: detections / availability
fit_detection <- ratiod(
  detections | available ~ effort + (1 | site),
  data = camera_data,
  family = ratiod_binomial()
)

# Extract detection probabilities
probs <- ratio(fit_detection)
```

### Count/Effort Ratios: `ratiod_poisson_gamma()`

When denominator is continuous effort:

```{r poisson-gamma-example}
# CPUE: catch count / fishing hours
fit_cpue <- ratiod(
  catch | hours ~ depth + season + (1 | vessel),
  data = fishing_data,
  family = ratiod_poisson_gamma()
)
```

## Formula Syntax

### Combined Formula (Recommended)

Use `num | denom ~ predictors` when both processes share the same predictors:

```{r combined-formula}
fit <- ratiod(
  catch | effort ~ depth + season + (1 | site),
  data = df,
  family = ratiod_poisson_gamma()
)
```

### Different Predictors per Process

Add process-specific terms with `formula_num` and `formula_denom`:

```{r different-predictors}
fit <- ratiod(
  catch | effort ~ (1 | site),           # Shared structure
  formula_num = ~ depth + season,         # Numerator-specific
  formula_denom = ~ weather,              # Denominator-specific
  data = df,
  family = ratiod_poisson_gamma()
)
```

### Separate Formulas

For complete control, provide `formula_denom` with a full formula:

```{r separate-formulas}
fit <- ratiod(
  catch ~ depth + season + (1 | site),
  formula_denom = effort ~ weather + (1 | site),
  data = df,
  family = ratiod_poisson_gamma()
)
```

## Shared Structure

### Default: Inferred from Matching Random Effects

When the same grouping variable appears in both formulas, ratiod shares the random effect by default:

```{r shared-default}
# site effect shared between catch and effort
fit <- ratiod(
  catch | effort ~ x + (1 | site),
  data = df,
  family = ratiod_poisson_gamma()
)
# Equivalent to: shared = ~ (1 | site)
```

### Explicit Shared Structure

```{r shared-explicit}
fit <- ratiod(
  catch | effort ~ x + (1 | site) + (1 | observer),
  shared = ~ (1 | site),  # Only site is shared
  data = df,
  family = ratiod_poisson_gamma()
)
```

### Independence (Use With Caution)

Request independent random effects:

```{r shared-independent}
fit <- ratiod(
  successes | trials ~ treatment + (1 | block),
  shared = ~ 0,  # Triggers warning about independence assumption
  data = experiment_data,
  family = ratiod_binomial()
)
```

## Spatial Structure

### CAR Model (Areal Data)

```{r spatial-car}
# Create adjacency matrix for sites
adj <- matrix(0, n_sites, n_sites)
# ... fill in neighbor relationships ...

# Fit with spatial structure
fit_spatial <- ratiod(
  count | effort ~ x + (1 | site),
  spatial = spatial_car(adj, level = "group", group_var = "site"),
  data = df,
  family = ratiod_poisson_gamma()
)
```

### BYM2 Model (Interpretable Spatial Fraction)

```{r spatial-bym2}
fit_bym2 <- ratiod(
  cases | population ~ age + (1 | region),
  spatial = spatial_bym2(adj, level = "group", group_var = "region"),
  data = epi_data,
  family = ratiod_binomial()
)
```

## Model Checking

### Posterior Predictive Checks

```{r pp-check}
# Density overlay
pp_check(fit, type = "dens_overlay", component = "numerator")
pp_check(fit, type = "dens_overlay", component = "denominator")

# Both components side by side
pp_check(fit, type = "dens_overlay", component = "both")

# Predictive intervals
pp_check(fit, type = "intervals")
```

### Model Comparison

```{r model-comparison}
# Compare nested models
fit1 <- ratiod(y | n ~ x, ...)
fit2 <- ratiod(y | n ~ x + z, ...)

# LOO comparison
ratiod_compare(fit1, fit2)

# Or compute LOO/WAIC individually
loo(fit1)
waic(fit1)
```

## Priors

### Default PC Priors

ratiod uses penalized complexity priors that favor simpler models:

```{r priors-default}
# Default: P(sigma > 1) = 0.01
fit <- ratiod(..., priors = ratiod_priors())
```

### Custom Priors

```{r priors-custom}
# More informative: expect smaller random effects
fit <- ratiod(
  ...,
  priors = ratiod_priors(
    sigma_U = 0.5,      # P(sigma > 0.5) = 0.01
    sigma_alpha = 0.01
  )
)

# Less informative: allow larger effects
fit <- ratiod(
  ...,
  priors = ratiod_priors(
    sigma_U = 2.0,
    sigma_alpha = 0.05  # P(sigma > 2) = 0.05
  )
)
```

## Quick Reference

### Main Functions

| Function | Purpose |
|----------|---------|
| `ratiod()` | Fit a ratiod model |
| `ratio()` | Extract ratio posteriors |
| `ratio_contrast()` | Compare ratios between groups |
| `pp_check()` | Posterior predictive checks |
| `loo()`, `waic()` | Model comparison criteria |
| `ratiod_compare()` | Compare multiple models |

### Families

| Family | Numerator | Denominator |
|--------|-----------|-------------|
| `ratiod_negbin_negbin()` | Overdispersed counts | Overdispersed counts |
| `ratiod_binomial()` | Successes | Trials (fixed) |
| `ratiod_poisson_gamma()` | Counts | Continuous effort |

### Spatial Functions

| Function | Type | Use Case |
|----------|------|----------|
| `spatial_car()` | ICAR | Simple areal smoothing |
| `spatial_bym2()` | BYM2 | Interpretable spatial fraction |

## See Also

- `vignette("philosophy")` - Why ratios are not data
- `?ratiod` - Main function documentation
- `?ratiod_priors` - Prior specification
- `?spatial_car` - Spatial structure
