title: "Spatial and Temporal Models"
author: "Gilles Colling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatial and Temporal Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = FALSE
)
```

## Overview

This vignette covers spatial and temporal extensions in ratiod. These allow borrowing strength across neighboring locations or time points, improving estimates in data-sparse regions.


# Spatial Models

ratiod supports two spatial priors for areal (discrete region) data:

| Function | Prior | Use Case |
|----------|-------|----------|
| `spatial_car()` | Intrinsic CAR (ICAR) | Simple spatial smoothing |
| `spatial_bym2()` | BYM2 | Interpretable spatial fraction |

Both require an adjacency matrix defining which regions are neighbors.

## Creating Adjacency Matrices

### From Coordinates (Grid Data)

```{r}
# Camera trap grid
n <- 25
coords <- expand.grid(x = 1:5, y = 1:5)

# Queen contiguity (8 neighbors)
adj <- matrix(0, n, n)
for (i in 1:n) {
  for (j in 1:n) {
    if (i != j) {
      dist <- sqrt((coords$x[i] - coords$x[j])^2 +
                   (coords$y[i] - coords$y[j])^2)
      if (dist <= sqrt(2) + 0.01) adj[i, j] <- 1
    }
  }
}
```

### From sf Objects

```{r}
library(sf)
library(spdep)

# From shapefile
regions <- st_read("regions.shp")
nb <- poly2nb(regions, queen = TRUE)
adj <- nb2mat(nb, style = "B", zero.policy = TRUE)
```

## ICAR Model: `spatial_car()`

The intrinsic conditional autoregressive (ICAR) prior encourages neighboring regions to have similar effects:

$$
\phi_i | \phi_{-i} \sim \text{Normal}\left(\frac{1}{n_i}\sum_{j \sim i} \phi_j, \frac{\sigma^2}{n_i}\right)
$$

where $j \sim i$ denotes neighbors of region $i$.

```{r}
fit <- ratiod(
  count | effort ~ x + (1 | region),
  spatial = spatial_car(
    adj = adj,
    level = "group",
    group_var = "region"
  ),
  data = df,
  family = ratiod_poisson_gamma()
)
```

### Parameters

- `adj`: Square binary adjacency matrix
- `level`: `"group"` (one effect per group) or `"obs"` (one per observation)
- `group_var`: Variable name matching the grouping factor (required if `level = "group"`)

## BYM2 Model: `spatial_bym2()`

The BYM2 reparameterization (Riebler et al., 2016) separates spatial structure from unstructured heterogeneity:

$$
\theta_i = \sigma \left(\sqrt{\rho} \cdot \phi_i^* + \sqrt{1-\rho} \cdot v_i\right)
$$

where:
- $\phi_i^*$ is the scaled ICAR component
- $v_i \sim \text{Normal}(0, 1)$ is unstructured noise
- $\rho \in [0, 1]$ is the proportion of variance that is spatial

```{r}
fit <- ratiod(
  cases | population ~ age + (1 | county),
  spatial = spatial_bym2(
    adj = adj,
    level = "group",
    group_var = "county"
  ),
  data = df,
  family = ratiod_binomial()
)
```

The $\rho$ parameter is directly interpretable: $\rho = 0.8$ means 80% of the random effect variance is spatially structured.

## Extracting Spatial Effects

```{r}
# Get posterior summaries
spatial_effects <- extract_spatial(fit)

# Map visualization
library(ggplot2)
library(sf)

regions$spatial_effect <- spatial_effects$mean
ggplot(regions) +
geom_sf(aes(fill = spatial_effect)) +
  scale_fill_gradient2()
```


# Temporal Models

ratiod supports temporal priors for time series structure:

| Function | Prior | Use Case |
|----------|-------|----------|
| `temporal_rw()` | Random walk | Smooth trends |
| `temporal_ar1()` | AR(1) | Mean-reverting dynamics |

## Random Walk: `temporal_rw()`

First-order random walk (RW1):

$$
\gamma_t | \gamma_{t-1} \sim \text{Normal}(\gamma_{t-1}, \sigma_\gamma^2)
$$

Second-order random walk (RW2) for smoother trends:

$$
\gamma_t | \gamma_{t-1}, \gamma_{t-2} \sim \text{Normal}(2\gamma_{t-1} - \gamma_{t-2}, \sigma_\gamma^2)
$$

```{r}
# First-order random walk
fit <- ratiod(
  count | effort ~ x + (1 | site),
  temporal = temporal_rw(
    time_var = "year",
    order = 1
  ),
  data = df,
  family = ratiod_poisson_gamma()
)

# Second-order for smoother trends
fit <- ratiod(
  count | effort ~ x,
  temporal = temporal_rw(
    time_var = "year",
    order = 2
  ),
  data = df,
  family = ratiod_poisson_gamma()
)
```

### Cyclic Random Walk

For seasonal patterns that wrap around (e.g., month 12 neighbors month 1):

```{r}
fit <- ratiod(
  count | effort ~ x,
  temporal = temporal_rw(
    time_var = "month",
    order = 1,
    cyclic = TRUE
  ),
  data = df,
  family = ratiod_poisson_gamma()
)
```

## AR(1): `temporal_ar1()`

Autoregressive order 1:

$$
\gamma_t = \rho \cdot \gamma_{t-1} + \epsilon_t, \quad \epsilon_t \sim \text{Normal}(0, \sigma_\gamma^2)
$$

The correlation parameter $\rho \in (-1, 1)$ controls persistence.

```{r}
fit <- ratiod(
  count | effort ~ x,
  temporal = temporal_ar1(
    time_var = "year"
  ),
  data = df,
  family = ratiod_poisson_gamma()
)
```

## Group-Specific Temporal Effects

By default, temporal effects are shared across all groups. For group-specific trends:

```{r}
fit <- ratiod(
  count | effort ~ (1 | site),
  temporal = temporal_rw(
    time_var = "year",
    group_var = "site",
    shared = FALSE
  ),
  data = df,
  family = ratiod_poisson_gamma()
)
```

## Extracting Temporal Effects

```{r}
# Year-level effects
temporal_effects <- extract_temporal(fit)

# Plot trend
library(ggplot2)
ggplot(temporal_effects, aes(x = year, y = mean)) +
  geom_ribbon(aes(ymin = q2.5, ymax = q97.5), alpha = 0.2) +
  geom_line() +
  labs(y = "Temporal effect")
```


# Combining Spatial and Temporal

For spatiotemporal data, include both:

```{r}
fit <- ratiod(
  count | effort ~ x + (1 | region),
  spatial = spatial_bym2(adj, level = "group", group_var = "region"),
  temporal = temporal_rw(time_var = "year"),
  data = df,
  family = ratiod_poisson_gamma()
)
```

This fits additive spatial and temporal effects. Full spatiotemporal interaction is planned for v1.2.


## Prior Specification

### Spatial Priors

```{r}
fit <- ratiod(
  ...,
  spatial = spatial_car(adj, level = "group", group_var = "region"),
  priors = ratiod_priors(
    tau_spatial_shape = 1.0,  # Gamma shape for precision
    tau_spatial_rate = 0.01   # Gamma rate for precision
  )
)
```

### Temporal Priors

```{r}
fit <- ratiod(
  ...,
  temporal = temporal_rw(time_var = "year"),
  priors = ratiod_priors(
    tau_temporal_shape = 2.0,
    tau_temporal_rate = 0.5
  )
)
```


## See Also

- `vignette("getting-started")` - Basic usage
- `vignette("workflows")` - Complete analysis examples
- `?spatial_car`, `?spatial_bym2` - Function documentation
- `?temporal_rw`, `?temporal_ar1` - Function documentation
